---
title: "Historic SIA Summary Analysis"
date: "`r Sys.Date()`"
# To generate: rmarkdown::render("scripts/response_sia_sweep/scripts/summarize_historic_sias.Rmd")
output:
  pdf_document:
    toc: false
    number_sections: true
header-includes:
  - \usepackage{float}
  - \floatplacement{figure}{H}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)

suppressPackageStartupMessages({
  library(tidyverse)
  library(jsonlite)
  library(dplyr)
  library(geosphere)
})
options(max.print = 1000)
```

# Introduction

This document summarizes the historic Supplementary Immunization Activities (SIAs) for Nigeria from 2017 onwards using vaccine types mOPV2, nOPV2, or tOPV. The analysis focuses on:
- Number of SIAs per year
- Time gaps between SIAs
- Geographic scope of SIAs (provinces and nodes)
- Geographic extent (distances between locations)


```{r load-data, echo = FALSE}
# Load the historic SIA schedule
historic_sias <- read_csv("../../../data/sia_historic_schedule.csv") %>%
  mutate(
    date = as.Date(date),
    year = year(date)
  )

# Filter to Nigeria
historic_sias <- historic_sias %>%
  filter(str_detect(dot_name, "NIGERIA")) %>%
  filter(year >= 2017) %>%
  filter(vaccinetype %in% c("mOPV2", "topv", "nOPV2"))

# Add lat/long to the historic_sias dataframe
node_lookup <- fromJSON("../../../data/node_lookup.json")
node_df <- bind_rows(node_lookup)
historic_sias <- historic_sias %>%
  left_join(node_df, by = c("dot_name" = "dot_name"))
```

# Number of SIAs Per Year

```{r sias-per-year, echo = FALSE}
sias_per_year <- historic_sias %>%
  group_by(year) %>%
  summarize(n_sias = n_distinct(date))
```

```{r sias-per-year-plot, fig.cap="Number of distinct SIA dates per year", echo = FALSE}
ggplot(sias_per_year, aes(x = year, y = n_sias, label = n_sias)) +
  geom_line() +
  geom_point() +
  geom_text(vjust = -0.5) +
  labs(
    x = "Year",
    y = "Number of SIAs"
  ) +
  theme_minimal()
```


```{r sias-per-year-summary, echo = FALSE}
cat("Summary of number of SIAs per year:\n")
summary(sias_per_year$n_sias)
```


# Time Between SIAs

```{r time-gaps, echo = FALSE}
avg_gaps <- historic_sias %>%
  arrange(dot_name, date) %>%
  group_by(dot_name) %>%
  mutate(
    gap_days = as.numeric(date - lag(date))
  )
```

```{r time-gaps-plot, fig.cap="Distribution of time between consecutive SIAs within districts", echo = FALSE}
ggplot(avg_gaps, aes(x = gap_days)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "white") +
  labs(
    x = "Time Between SIAs (days)",
    y = "Count"
  ) +
  theme_minimal()

cat("Summary of gap between SIAs (days):\n")
summary(avg_gaps$gap_days)
```

# Geographic Scope of SIAs

## Provinces per SIA

```{r provinces-per-sia, echo = FALSE}
avg_provinces_per_sias <- historic_sias %>%
  group_by(date) %>%
  summarize(n_provinces = n_distinct(adm1))
```

```{r provinces-per-sia-plot, fig.cap="Distribution of the number of provinces covered per SIA", echo = FALSE}
ggplot(avg_provinces_per_sias, aes(x = n_provinces)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(
    x = "Number of Provinces per SIA",
    y = "Count"
  ) +
  theme_minimal()

cat("Summary of provinces per SIA:\n")
summary(avg_provinces_per_sias$n_provinces)
```

## Nodes per SIA

```{r nodes-per-sia, echo = FALSE}
avg_nodes_per_sias <- historic_sias %>%
  group_by(date) %>%
  summarize(n_nodes = n())
```

```{r nodes-per-year, echo = FALSE}
avg_nodes_per_sias_per_year <- historic_sias %>%
  group_by(year, date) %>%
  group_by(year) %>%
  summarize(n_nodes = n())

# print(avg_nodes_per_sias_per_year)
```

```{r nodes-per-year-plot, fig.cap="Total nodes covered by SIAs per year", echo = FALSE}
ggplot(avg_nodes_per_sias_per_year, aes(x = year, y = n_nodes, label = n_nodes)) +
  geom_line() +
  geom_point() +
  geom_text(vjust = -0.5) +
  labs(
    x = "Year",
    y = "Number of Nodes per SIA"
  ) +
  theme_minimal()

cat("Summary of nodes per SIA:\n")
summary(avg_nodes_per_sias$n_nodes)
```


\newpage
# Geographic Extent of SIAs

Calculate pairwise distances between all locations within each SIA to understand the geographic spread.

```{r distance-function, echo = FALSE}
# Function to calculate pairwise distances for a group
calc_pairwise_distances <- function(df) {
  n <- nrow(df)
  if (n < 2) {
    return(data.frame(
      dot_name_1 = character(),
      dot_name_2 = character(),
      distance_km = numeric()
    ))
  }

  # Create all pairs
  pairs <- expand.grid(i = 1:n, j = 1:n) %>%
    filter(i < j)

  # Calculate distances
  results <- data.frame(
    dot_name_1 = df$dot_name[pairs$i],
    dot_name_2 = df$dot_name[pairs$j],
    distance_km = distHaversine(
      cbind(df$lon[pairs$i], df$lat[pairs$i]),
      cbind(df$lon[pairs$j], df$lat[pairs$j])
    ) / 1000
  )

  return(results)
}
```

```{r calculate-distances, echo = FALSE}
# Apply to each date
distance_summary <- historic_sias %>%
  group_by(date) %>%
  group_modify(~ calc_pairwise_distances(.x)) %>%
  ungroup() %>%
  group_by(date) %>%
  summarise(
    n_locations = n_distinct(c(dot_name_1, dot_name_2)),
    n_pairs = n(),
    mean_dist_km = mean(distance_km),
    max_dist_km = max(distance_km),
    min_dist_km = min(distance_km)
  )
```

```{r distance-plot, fig.cap="Distribution of maximum pairwise distance within each SIA", echo = FALSE}
ggplot(distance_summary, aes(x = max_dist_km)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  labs(
    x = "Maximum Distance Between Locations in an SIA (km)",
    y = "Count"
  ) +
  theme_minimal()

cat("Summary of maximum distance (km) within SIAs:\n")
summary(distance_summary$max_dist_km)
```
